---
title: "Analysis of Cumulative COVID-19 Cases across Nations"
author: "Ngoc Duong, Jared Garfinkel, Adeline Shin, Alyssa Vanderbeek"
date: "1 May 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(directlabels)
library(countrycode)

load("./curves.RData")
```

## Introduction

COVID-19 has recently caused a worldwide pandemic. While the illness began in Wuhan, China, it quickly spread to other countries, and the world saw an exponential rise in both cases and deaths. Modeling the trajectories of this growth at the community level - country, state, city, etc. - is important for influencing policy in the effort to curb this pandemic. In this project, we use an exponential model to estimate and predict country-level COVID-19 cases. We also sought to understand whether these countries can be grouped - comparing two clustering algorithms - by the features of their trajectories, and what those influencing features might be. 


## Methods

### Data

We used data from Johns Hopkins, which holds information about the daily total number of cases in a given region over a 3-month period, between January 22 and April 20. Information in the data includes the country or region, the city, latitude and longitude coordinates, and the cumulative number of confirmed cases and deaths for that day. 

We are interested primarily in the virus trajectory over time. We looked at the country level to model the cumulative number of cases over time, beginning on the day that each country reported its first COVID-19 case.


### Part 1: Fitting the models

We fit exponential models to each country in our data using coordinate-wise optimization over a pre-specified interval to minimize our loss function, defined here as the mean sum of squares (SS): $$loss = \frac{\sum_{i = 1}^{T_j} (\hat{X_i} - X_i)}{T_j}$$ for the $j^{th}$ country and $i^{th}$ time point. The exponential distribution is modeled as $$f(t) = \frac{a}{1 + exp\{-b(t - c)\}}$$, where $t$ represents the number of days since the first documented infection, $a$ is the maximum value or the curve, $b$ is the growth rate, and $c$ is the midpoint (the time at which the growth rate $b$ is highest. 

To assess the prediction ability of the country-specific models, we predicted the total number of confirmed cases for the following 8 days, from April 21-29, and looked at the RMSE of the predictions.


### Part 2: Clustering

We clustered the data with both K-means and Gaussian mixture model. For both methods, we first standardize the data as follows: 
$$\text{standardized}(x_{i}) = \frac{x_{i} - \bar{x}}{sd(x)}$$ for $i = 1,2,..., n$

**K-means**

First, we produce a set of initial centroids (sample means) $\mu_1^{(0)}, \mu_2^{(0)},...,\mu_k^{(0)}\}$. The k-means algorithm is based on that initialized setup to iterate the following two steps:

*1) Find optimal cluster assignment given fixed centroids* 

For each data point $x_i$, compute $d_i$, the Euclidean distance between $x_i$ and the centroids, then assign $x_i$ to cluster with the smallest $d_i$. This is equivalent to: 

$$r_{i,j}^{(v+1)} = I\{j=\arg\min_j \|\mathbf x_i - \mu_j^{(v)}\|\}$$

*2) Calculate cluster centers using the cluster assignment in the previous step*

This steps involves minimizing $J(r, \mu)$ over $\mu$: 

$$\mu_j^{(v+1)} =\frac{\sum_{i=1}^n\mathbf x_i r_{i,j}^{(v+1)}}{\sum_{i=1}^n r_{i,j}^{(v+1)}} $$

In other words, we update the new cluster means by taking the average of all datapoints that belong to that cluster using a weighted probability distribution where point $x_i$ is chosen. This reflects the new cluster assignments of the datapoints.

Then, we repeat these two steps until k centers have been chosen/converged.

**Gaussian mixtures**

Let $\{x_1,x_2,...,x_n\} \in R^3$ be a collection of 3-dimensional datapoints. Assuming they come from underlying Gaussian distributions, we can implement the EM algorithm to identify what distribution each points belongs to, which consists of three steps: initialization, E-step, and M-step. In short, the algorithm randomly selects k different clusters and starts with finding weights, mean, and covariance matrix for each cluster. Then, E-step can be applied on those values. These two steps are then reiteratively applied back-and-forth until centroids converge or maximum iteration threshold is reached.

*Initialization*

Gaussian mixture models assume that each latent class has different means and covariances. However, since each class is unknown, we begin by intializing the parameters and iteratively updating. Here, we chose to initialize 3 (random) different classes.

Then, we initialize a weight (posterior probability) matrix $\gamma$ with size, where each row contains the mean value for each of the three Gaussians. Then, the values for covariance matrix list, where each element in the list represents a covariance matrix for each Gaussian. Finally, we can randomly assign data to each Gaussian with the weight matrix, where each column shows the probability each point belongs to each of the three clusters.

*E-step*

In E-step, each observation is assigned a weight (responsibility) for each cluster, based on the likelihood of each of the corresponding Gaussian using the initialized parameters (means and covariance matrix and the function dmvnorm). The responsibility is defined as:

$$\hat\gamma_{i, k} ^{(t)}= P(r_{i,k}=1 |\mathbf x_i,  \theta^{(t)}) =  
\frac{p_k^{(t)}f(\mathbf x_i|\boldsymbol \mu_k^{(t)}, \Sigma_k^{(t)})}
{\sum_{j=1}^K f(\mathbf x_i|\boldsymbol \mu_j^{(t)}, \Sigma_j^{(t)})}$$

*M-step*

In the M-step, we aim to maximize the complete log-likehood function and thus revising the parameters. In this step, each observation contributes to the weighted means and covariances for every cluster. Here, the mixing component is $p_k = \frac{n_k}{n} (1)$, also a prior for each class estimated from the data. 
Let $n_k = \sum_i^n{\gamma_{ik}} (2)$, we have 
$$\mu_k = \frac{1}{m_k} \sum_i{\gamma_{ik}x_i} (3)$$, and 
$$\Sigma_k = \frac{1}{n_k} \sum_i{\gamma_{ik}(x_i – \mu_k)^T(x_i-\mu_k)} (4)$$


We start by updating the mixture components (i.e. prior probabilities). Each datapoint contributes to a $\gamma_{i,k}$ which corresponds to one column vector each class. Given the available information, using (1) and (2), we can update the prior probabilities. Following the formulas (3) and (4), we can update $\mu_k^{t+1}$ and $\Sigma_k^{t+1}$ as our new Gaussian density parameters. This process involves standard calculation of mean and variance for a Gaussian distribution. However, during this process, we also use weights $\gamma_{ik}$ and $\frac{1}{m_k}$ to weight each point in our mean and covariance estimation.

Once the likelihood has been maximized in this step, we loop back and iterate the process until a maximum is found or an iteration threshold has been reached.



## Results

### Data and model fitting

There are 186 countries in the dataset. Generally speaking, countries that were hit with the virus earlier have a higher total number of confirmed cases (Figure 1).

```{r dist, fig.height = 8, fig.width = 8}
total_cases = by_country %>%
  group_by(country_region) %>%
  slice(n()) %>%
  rowwise() %>%
  mutate(gr = if_else(confirmed_cases >= 1e4 & confirmed_cases < 1e5, "10,000 - 99,999", 
                      if_else(confirmed_cases >= 5e3 & confirmed_cases < 1e4, "5,000 - 9,999",
                              if_else(confirmed_cases >= 1e3 & confirmed_cases < 5e3, "1,000 - 4,999",
                                      if_else(confirmed_cases >= 500 & confirmed_cases < 1e3, "500 - 999",
                                              if_else(confirmed_cases >= 100 & confirmed_cases < 500, "100 - 499", 
                                                      if_else(confirmed_cases < 100, "0 - 99", "100,000 +"))))))) %>%
  mutate(gr = factor(gr, level = c("0 - 99",
                                   "100 - 499",
                                   "500 - 999",
                                   "1,000 - 4,999",
                                   "5,000 - 9,999",
                                   "10,000 - 99,999",
                                   "100,000 +")))


totals_count = table(total_cases$gr) %>%
  as.data.frame() %>%
  mutate(y = c(15, 6, 4, 5, 3, 3, 2),
         x = 15,
         text = paste("n = ", Freq)) %>%
  dplyr::rename(., gr = Var1)


total_cases %>%
  ggplot(aes(x = t)) +
  geom_histogram() +
  geom_text(data = totals_count,
            aes(x = x, y = y, label = text),
            fontface = "bold",
            hjust = 0,
            size = 5) +
  facet_grid(gr ~ ., scales = "free") + 
  labs(x = "Days since first infection",
       y = "Number of countries",
       title = "Number of days since first infection") +
  theme_bw() + 
  theme(strip.text.y = element_text(size = 10, angle = 270))
```


Since the midpoint milestone, the cumulative number of cases per day have been rising at a decreasing rate. There are 93 countries that are past the midpoint, or peak, of COVID-19 spread. Of these, 63 are at least a week past, 38 are at least 2 weeks past, and 3 countries (China, South Korea, and Cambodia) are at least 1 month past the peak.

It appears that the countries with the highest growth rate (b) have comparably low expected total number of cases. Table 1 gives this information for the top 10 largest growth rate countries, all of which have passed their midpoint. 

```{r}
summary_country_df = by_country %>% 
  rename(region = country_region) %>%
  left_join(., param_df1, by = "region")  

peak_data = summary_country_df %>% group_by(region) %>% 
  mutate(max_t = max(t)) %>% 
  dplyr::select(-t, -confirmed_cases, -fatalities) %>% 
  distinct(region, .keep_all = TRUE) 

peak_data %>% ungroup() %>% top_n(10, b) %>% 
  dplyr::arrange(desc(b)) %>%
  mutate(c = round(c, 0)) %>%
  knitr::kable(caption = "Top 10 countries with high growth rate",
               col.names = c("Country",
                             "a",
                             "b",
                             "c",
                             "Days obs. since first infection (t)"))
```


### Model fit

A visual inspection suggests that the exponential distribution is an appropriate model for the data across countries. Figure 2 shows the observed and fitted data for select countries.

```{r highlighted_fits}

```

Using the estimated parameter for the period until 04/20/2020, we were able to make predictions for the following 8 days until 04/29/2020. The median RMSE for these new predictions across all countries is 131.97. The lowest RMSE is 0.94, and mean RMSE is within acceptable range given the rapid surge in recent number of cases. The maximum RMSE, however, is quite high, at more than 100,000. 

### Clustering

K-means and GMM both give interpretable groupings.

Under a K-mean algorithm with 4 clusters, all three components of the exponential model contributed to the groupings (Figure 3).

- Group 1: Moderate total expected cases, moderate growth rate, short time to peak.
- Group 2: High total expected cases, moderate growth rate, long time to peak.
- Group 3: Low-moderate total expected cases, high growth rate, short time to peak.
- Group 4: Wide range in total expected case, low growth rate, long time to peak.

```{r}

```

Under a GMM with 3 clusters, the main discriminating feature is the total expected number of cases (Figure 4). In this case, we can observe low-, moderate-, and high-case level groups.

```{r}

```


## Discussion


This is the case of the US. As we did not model each state separately but instead lump all data under the country USA, it is reasonable to see high deviances between the number of predicted cases and actual confirmed cases. As we all know, different states within the US are seeing the number of confirmed cases rising at different rates, on top of discrepancies in testing availability throughout the country, there is no one-size-fits-all model that might capture precisely these underlying trends. 

All in all, besides the high RMSE in the US, which is due to vast differences between regions populations and state policies regarding public health interventions, the RMSE for all other countries are within reasonable range, which gives us confidence in the model's ability to predict well future cases. 


Despite the relative similarity in the decision boundaries of the two clustering methods, Gaussian mixtures model produced a somewhat smoother decision boundary/plane. (3D plot?)

We also might want to take into account some considerations:
Gaussian mixture models converge well when the class densities are better separated, so this also depends on how fine the parameter estimates are. 




\pagebreak
## R functions

```{r fit_curves, eval = FALSE, echo = TRUE}
# calculate predictions given data and parameter values
logistic_pred = function(t, a, b, c) {
  t  # predictor: days since first infection
  y_hat = a / (1 + exp(-b*(t - c)))
  return(y_hat)
}

## Our goal is to minimize the loss function (SSE)
loss = function(data, a, b, c) {
  y = data$confirmed_cases
  y_hat = logistic_pred(data$t, a, b, c) # predicted 
  error = y_hat - y
  
  loss = sum( error^2 ) / length(y)
  return(loss)
}


## Coordinate-wise optimization to fit logistic curve to data
fit_curve = function(data, aseq, bseq, cseq) {
  c_lim = c(min(cseq), max(cseq))
  
  # matrix of optimal values of c across all combos of a and b values
  cvals = sapply(aseq, function(a){
    sapply(bseq, function(b){
      optimize(loss, interval = c_lim, data = data, a = a, b = b)$minimum
    })
  })
  # matrix of the computed loss for all combos
  closs = sapply(aseq, function(a){
    sapply(bseq, function(b){
      optimize(loss, interval = c_lim, data = data, a = a, b = b)$objective
    })
  })
  
  loc = which(closs == min(closs), arr.ind = TRUE)
  a = aseq[loc[2]]
  b = bseq[loc[1]]
  c = cvals[loc[1], loc[2]]
  
  return(cbind(a = a, b = b, c = c))
}
```


```{r kmeans, eval = FALSE, echo = TRUE}
#partition of data such that squared error between empirical mean and points in each cluster/partition is minimized
km_func <- function(data, k){
  p <- ncol(data)  # number of parameters
  n <- nrow(data)  # number of observations
  diff = 1
  iter = 0
  itermax = 50
  while(diff > 1e-4 && iter <= itermax){
    #initial centroids
    if(iter == 0){
      centroid = data[sample(nrow(data), k),]
      centroid_mem = centroid
    }

    #assign to cluster 
    d = sapply(1:k, function(c) sapply(1:n, function(i) {sum((centroid[c,] - data[i,])^2)}))
    
    cluster = apply(d, 1, which.min)
    
    #recalculate cluster 
    centroid = t(sapply(1:k, 
                         function(c) {apply(data[cluster == c,], 2, mean)}
                         ))
    
    #recalculate distance
    diff = sum((centroid - centroid_mem)^2)
    iter = iter + 1 
    centroid_mem = centroid
  }
  return(list(centroid = centroid, cluster = cluster))
}
```


```{r gmm, eval = FALSE, echo = TRUE}
gmm_func <- function(X, k){
  #setting
  data <- as.matrix(X) 
  #%>% scale()
  N <- nrow(data)
  q <- ncol(data)
  p_j <- rep(1/k, k)
  mu <-  data[sample(N, k),] %>% as.matrix()   
  covmat <- diag(ncol(data))
  
  covList <- list()
  for(i in 1:k){
    covList[[i]] <- covmat
  }
 
  count=1
  while(count <100){
    mu0 <- mu
    
    # E-step: Evaluate posterior probability, gamma
    gamma <- c()
    for(j in 1:k){
      gamma2 <- apply(data,1, mvtnorm::dmvnorm, mu[j,], covList[[j]])
      gamma <- cbind(gamma, gamma2)
    }
    
    # M- step: Calculate mu
    tempmat <- matrix(rep(p_j,N),nrow=N,byrow = T)
    r <- (gamma * tempmat) / rowSums(gamma*tempmat)  
    mu <- t(r) %*% data / colSums(r) 
    
    # M- step: Calculate Sigma and p
    for(j in 1:k){
      sigma <- matrix(rep(0,q^2),ncol=q)
      for(i in 1:N){
        sigma = sigma + r[i,j] * (data[i,]-mu0[j,]) %*% t(data[i,]-mu0[j,])   
      }
      covList[[j]] <- sigma/sum(r[,j])
    }
    p_j <- colSums(r)/N
    count = count + 1
  }
  
  cluster <- which(r == apply(r, 1, max), arr.ind = T)
  cluster <- cluster[order(cluster[,1]),]
  return(list(cluster = cluster))
}
```

